Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-02-05 article/04_Discussion.Rmd
some love for the conclusion

The **A** sections of the figures \ref{development_burial_type} and \ref{development_burial_construction} with absolute grave numbers per region and year indicate areas and periods with a lot and others with few points of data. For some there is not enough information to make any decision about the relative dominance and for many the general relevance and accuracy of the resulting time series can be questioned. There is a variety of reasons for this. These should also be considered for future data acquisition in Radon-B and beyond.

@@

This paper is based on ^14^C dates which were prepared for the dating of graves. There is a fundamental break between that great, theoretical archaeology, which wants to develop comprehensive narratives and only pay quantitative attention to the individual case, and the on-site research or emergency excavation, which has to dedicate a measure of respect and sometimes a lot of time to each grave. In this thesis, ^14^C data are converted into a proxy of calibrated density distributions within artificial working regions, which is completely detached from the individual context. ^14^C data on graves are used to attribute a feature to the absolute chronological context or to better describe the relative development within a site. The questions underlying the creation of a ^14^C date are therefore not equivalent to those imposed in this paper.

>>>>

The **A** sections of the figures \ref{development_burial_type} and \ref{development_burial_construction} with absolute grave numbers per region and year indicate areas and periods with a lot and others with few points of data. For some there is not enough information to make any decision about the relative dominance and for many the general relevance and accuracy of the resulting time series can be questioned. There are a variety of reasons for this. These should also be considered for future data acquisition in Radon-B and beyond.

@@

This paper is based on ^14^C dates which were prepared for the dating of graves. There is a fundamental break between that great, theoretical archaeology, which wants to develop comprehensive narratives and only pay quantitative attention to the individual case, and the on-site research or emergency excavation, which has to dedicate a measure of respect and sometimes a lot of time to each grave. In this paper, ^14^C data are converted into a proxy of calibrated density distributions within artificial working regions, which is completely detached from the individual context. ^14^C data on graves are used to attribute a feature to the absolute chronological context or to better describe the relative development within a site. The questions underlying the creation of a ^14^C date are therefore not equivalent to those imposed in this paper.

### 2019-02-05 article/05_Conclusion.Rmd
some love for the conclusion

The major questions posed in the introduction to this paper have been answered : *Bulk radiocarbon data is a powerful source to trace general trends in prehistoric Europe* and *allows long-term observations concerning cultural distance*. Radiocarbon dates can be transformed to meaningful regional time series, if they can be linked to the relevant meta information. Although only `r r("gpr size", sf_desc)` graves of a maybe six-figure statistical population that can be estimated for all burials in the European Bronze Age [@nikulkaArchaologischeDemographieMethoden2016] have been available for this analysis, the resulting reconstruction of general burial tradition trends in wast areas of Central, Northern and North-western Europe comes close to the one achieved by literature research. The time series can be employed for cultural distance calculation. They yield the surprising observation that cultural distance and spatial distance do not significantly correlate. Although this interpretation must be treated with caution, it seems as if the spread of burial traditions are not simple *Diffusion of Innovation* events as introduced by @rogersDiffusionInnovations1962, but the result of more complex interactions between competing ideas.

Future research is required to de-tangle these interactions -- of course on the basis of more precise data. This endeavour will not end with the simple description of social learning cascades, but with the analysis of the initial reasons for diversification. From the point of view of Cultural Complexity -- the question, which factors accelerate or decelerate the variability of human behaviour [@henrichDemographyCulturalEvolution2004; @mesoudiWhatCumulativeCultural2018; @vaesenPopulationSizeDoes2016] -- the long term dominance fluctuations of one burial rite over the other are remarkable. Its standard explanation models (among others population density and environmental risk) can hardly be applied to stylistic variables.

@@

1. For this paper burial traditions were the main object of research, but the same methodological approach could be applied to reconstruct other aspects of cultural behaviour. The idea to systematically collect data points that combine spatial and temporal location with simplified meta information about cultural phenomena might be a way to achieve better reproducibility of overarching narratives. Fast and easy *reproducibility* is of crucial importance to achieve scientific progress and it includes the notion of rather technical *preproducibility* [@starkReproducibilityMustCome2018]: Data (and code) must be shared to fuel advances in anthropological theory building [@kitzesPracticeReproducibleResearch2017; @marwickComputationalReproducibilityArchaeological2017; @marwickPackagingData2018].

2. This paper is built upon Cultural evolution theory (see section \ref{introduction}). It not just relies on terminology and methodology developed in this field, the process of abstraction necessary to go from radiocarbon dates to cultural distance was also heavily influenced by Evolutionary archaeology. This theoretical framework might be a source of inspiration and methods for other projects. The adoption and adaption of statistical tools and models from population genetics into anthropology was and is a unique characteristic of Cultural Evolution theory [@cavalli-sforzaCulturalTransmissionEvolution1981; @boydCultureEvolutionaryProcess1985; @teltserEvolutionaryArchaeologyMethodological1995; @obrienEvolutionaryArchaeologyTheory1996]. On the other hand, archaeology is explicitly concerned with human behaviour over big areas and long time scales and can offer a unique perspective to other anthropological fields [@garveyCurrentPotentialRoles2018].

>>>>

*The main trends in the distribution of burial rites in Bronze Age Europe can be detected in bulk radiocarbon data* and indeed *they allow long-term observations concerning cultural distance*. Radiocarbon dates can be transformed to meaningful regional time series, if they can be linked to the relevant meta information. Although only `r r("gpr size", sf_desc)` graves -- a small fraction of maybe hundreds of thousands burials that can be estimated for the European Bronze Age [@nikulkaArchaologischeDemographieMethoden2016] -- have been available for this analysis, the resulting reconstruction of general burial tradition trends in wast areas of Central, Northern and North-western Europe comes close to the one achieved by literature research. The time series can be employed for cultural distance calculation. They yield the surprising observation that cultural distance and spatial distance do not significantly correlate. Although this interpretation must be treated with caution, it seems as if the spread of burial traditions are not simple *Diffusion of Innovation* events as introduced by @rogersDiffusionInnovations1962, but the result of more complex interactions between competing ideas.

Future research is required to untangle these interactions -- of course on the basis of more precise data. This endeavour will not end with the simple description of social learning cascades, but with the analysis of the initial reasons for diversification. From the point of view of Cultural complexity -- the question, which factors accelerate or decelerate the variability of human behaviour [@henrichDemographyCulturalEvolution2004; @mesoudiWhatCumulativeCultural2018; @vaesenPopulationSizeDoes2016] -- the long term dominance fluctuations of one burial rite over the other are remarkable. Its standard explanation models (among others population density and environmental risk) can hardly be applied to stylistic variables.

@@

1. For this paper burial traditions were the main object of research, but the same approach could be applied to reconstruct other aspects of cultural behaviour. The idea to systematically collect data points that combine spatial and temporal location with simplified meta information about cultural phenomena might be a way to achieve better reproducibility of overarching narratives. Fast and easy *reproducibility* is of crucial importance to achieve scientific progress and it includes the notion of rather technical *preproducibility* [@starkReproducibilityMustCome2018]: Data (and code) must be shared to fuel advances in anthropological theory building [@kitzesPracticeReproducibleResearch2017; @marwickComputationalReproducibilityArchaeological2017; @marwickPackagingData2018].

2. This paper is built upon Cultural transmission and thereby Cultural evolution theory (see section \ref{introduction}). It not just relies on terminology and methodology developed in this field, the process of abstraction necessary to go from radiocarbon dates to cultural distance was also heavily influenced by Evolutionary archaeology. This theoretical framework might be a source of inspiration and methods for other projects. The adoption and adaption of statistical tools and models from population genetics into anthropology was and is a unique characteristic of Cultural Evolution theory [@cavalli-sforzaCulturalTransmissionEvolution1981; @boydCultureEvolutionaryProcess1985; @teltserEvolutionaryArchaeologyMethodological1995; @obrienEvolutionaryArchaeologyTheory1996]. On the other hand, archaeology is explicitly concerned with human behaviour over big areas and long time scales and can reciprocate with a unique perspective [@garveyCurrentPotentialRoles2018].
